A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2017-01-19 doc/paper1-submission/paper5.tex
minor tweaks

\begin{abstract}
We present a new supertree method that enables rapid estimation of a
summary tree on the scale of millions of leaves.  This supertree
method summarizes a collection of input phylogenies and an input taxonomy.
We introduce formal goals and criteria for such a supertree to satisfy
in order to transparently and justifiably represent the input
trees. In addition to producing a supertree, our method computes
annotations that describe which grouping in the input trees support
and conflict with each group in the supertree.

@@

We compare our supertree construction method to a previously published
supertree construction method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2,760 to 1,357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

@@

The Open Tree of Life project seeks to build a platform for summarizing
what is known about phylogenetic relationships across all of Life
\citep{HinchliffEtAl2015}. One primary goal of the project is to
build a summary tree from a comprehensive taxonomic tree and a set
of published trees. The summary tree is intended to transparently
and justifiably represent phylogenetic information from these inputs.
The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter,
publication in preparation). The phylogenetic inputs are published
trees that have been curated to align the tips to OTT and to identify
the correct rooting (see \citealt{McTavishEtAt2015} for further details
of the curation tools). Unlike OTT, these phylogenetic trees do not
include all leaf taxa. The inputs (taxonomy and phylogenetic trees)
and the output summary supertree are all rooted. Here we describe
the software pipeline (\emph{propinquity}) that summarizes and integrates
these smaller source trees and the taxonomy tree into a single supertree
and the noteworthy tools for manipulating and solving supertrees in
the \emph{otcetera} package.

@@

For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
are empty\footnote{The intersection of the right sides of the splits is considered to
be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
sides implicitly contain the root.}. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
the shorthand of referring to an edge $k$ when we mean the split
$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
we mean that $S(j)$ conflicts with $S(k)$.

@@

An appealing goal for the summarization would be to find the supertree
that displays the largest number of input tree edges. As discussed
in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
problem is known to be $NP$-hard via a reduction from Max-Clique \citep{Karp1972}.
In addition to being computationally daunting, this formulation of
the supertree problem does not provide biologists who use the summarization
tool with an obvious avenue for fixing perceived problems with the
summary tree. For example, a grouping that a biologist expected may
not be present in the supertree, but it may not conflict with any
of the input groupings which are displayed. This can happen because
displaying both node $a$ from $T_{1}$ and node $b$ from tree $T_{2}$
in a summary tree may only be possible by displaying a grouping that
is present in no input tree. All other factors being equal, if this
implied grouping conflicts with input node $c$ in tree $T_{3}$,
then $c$ will not be displayed in the summary tree, but a biologist
will not necessarily know how to fix this problem. One solution is
to use a ranking of groupings. If an expert were quite confident in
the $c$ grouping, then she could assign that input node a high ranking.
A supertree that used ranks could then recover this grouping even
if its inclusion did not increase the total number of input nodes
that are displayed by the summary tree. Figure (\ref{fig:pairwisecompat})
shows an example of 3 input trees for which there is no pairwise incompatibility,
but no solution displays all of the input groups. Alternative rankings
of inputs can result in one of three summary trees shown in panels
(d-f).

@@

So, we also consider MRP and MinCut.  MRP can display edges not found
in any input tree, and so fails criterion #1.  MRP is used in many
studies that attempt to infer the true tree from a lot of input
trees.  To some extent, MRP tries to treat each tree as data, which
differs from our approach here.

@@

A number of other supertree methods exist, including Matrix
Representation with Parsimony (MRP) and MinCut.  MRP
% http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/
MRP - displays unsupported groups (#1), not rank-based (#2), ?
SuperFine - computes a strict consensus, then MRP
Minimum Flip - 
(Modified) MinCut -

>>>>

\begin{abstract}
We present a new supertree method that enables rapid construction of a
summary tree on the scale of millions of leaves.  This supertree
method summarizes a collection of input phylogenies and an input taxonomy.
We introduce formal goals and criteria for such a supertree to satisfy
in order to transparently and justifiably represent the input
trees. In addition to producing a supertree, our method computes
annotations that describe which grouping in the input trees support
and conflict with each group in the supertree.

@@

We compare our supertree construction method to a previously published
method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2,760 to 1,357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

@@

The Open Tree of Life project seeks to build a platform for summarizing
what is known about phylogenetic relationships across all of Life
\citep{HinchliffEtAl2015}. One primary goal of the project is to
build a summary tree from a comprehensive taxonomic tree and a set
of published trees. The summary tree is intended to transparently
and justifiably represent phylogenetic information from these inputs.
The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter;
publication in preparation). The phylogenetic inputs are published
trees that have been curated to align the tips to OTT and to identify
the correct rooting (see \citealt{McTavishEtAt2015} for further details
of the curation tools). Unlike OTT, these phylogenetic trees do not
include all leaf taxa. The inputs (taxonomy and phylogenetic trees)
and the output summary supertree are all rooted. Here we describe
the software pipeline (\emph{propinquity}) that summarizes and integrates
these smaller source trees and the taxonomy tree into a single supertree
and the noteworthy tools for manipulating and solving supertrees in
the \emph{otcetera} package.

@@

For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
are empty.
Note that the intersection of the right sides of the splits is considered to
be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
sides implicitly contain the root. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
the shorthand of referring to an edge $k$ when we mean the split
$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
we mean that $S(j)$ conflicts with $S(k)$.

@@

An appealing goal for the summarization would be to find the supertree
that displays the largest number of input tree edges. As discussed
in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
In addition to being computationally daunting, this formulation of
the supertree problem does not provide biologists who use the summarization
tool with an obvious avenue for fixing perceived problems with the
summary tree. For example, a grouping that a biologist expected may
not be present in the supertree, but it may not conflict with any
of the input groupings which are displayed. This can happen because
displaying both node $a$ from $T_{1}$ and node $b$ from tree $T_{2}$
in a summary tree may only be possible by displaying a grouping that
is present in no input tree. All other factors being equal, if this
implied grouping conflicts with input node $c$ in tree $T_{3}$,
then $c$ will not be displayed in the summary tree, but a biologist
will not necessarily know how to fix this problem. One solution is
to use a ranking of groupings. If an expert were quite confident in
the $c$ grouping, then she could assign that input node a high ranking.
A supertree that used ranks could then recover this grouping even
if its inclusion did not increase the total number of input nodes
that are displayed by the summary tree. Figure (\ref{fig:pairwisecompat})
shows an example of 3 input trees for which there is no pairwise incompatibility,
but no solution displays all of the input groups. Alternative rankings
of inputs can result in one of three summary trees shown in panels
(d-f).

@@

So, we also consider MRP and MinCut.  MRP can display edges not found
in any input tree, and so fails criterion \#1.  MRP is used in many
studies that attempt to infer the true tree from a lot of input
trees.  To some extent, MRP tries to treat each tree as data, which
differs from our approach here.

@@

A number of other supertree methods exist, including Matrix
Representation with Parsimony (MRP) and MinCut.  MRP
% http://currents.plos.org/treeoflife/article/reweaving-the-tapestry-a-supertree-of-birds/
MRP - displays unsupported groups (\#1), not rank-based (\#2), ?
SuperFine - computes a strict consensus, then MRP
Minimum Flip - 
(Modified) MinCut -
